Spatial variation in allometric growth of invasive lionfish has management implications

### 2019-02-11 docs/Manuscript.Rmd
SF final edit

---
title: "Spatial variation in allometric growth of invasive lionfish has management implications"
author: "Juan Carlos Villaseñor-Derbez, Sean P. Fitzgerald"
institution: "Bren School of Environmental Sciences and Management, University of California Santa Barbara, Santa Barbara, California, U.S."
abstract: |
  Lionfish (\textit{Pterois volitans / miles}) are an invasive species in the Western Atlantic and the Caribbean. Improving management of invasive lionfish populations requires accurate total biomass estimates, which depend on accurate estimates of allometric growth, but sedentary species like lionfish often exhibit high levels of spatial variation in life history characteristics. We reviewed 17 published length-weight relationships for lionfish taken throughout their invasive range and found regional differences that led to significant misestimates when calculating weight from length observations. The spatial pattern we observed is consistent with findings from other studies focused on genetics or length-at-age. We show that the use of \textit{ex situ} parameters can result in up to a threefold under- or overestimation of total weight for an individual organism. When estimating total biomass, the use of other parameters may produce estimates between 76.2% and 140% of true observed biomass. These findings can have implications for management in terms of predicting effects on local ecosystems, evaluating the effectiveness of removal programs, or estimating biomass available for harvest.
keywords:
  - Lionfish
  - invasive species
  - length-weight
  - allometric growth
  - regional variations
bibliography: references.bib
output:
  pdf_document:
    keep_tex: yes
    number_sections: no
    template: peerj_template.tex
---

@@

The feasibility of establishing fisheries through lionfish removal programs has been extensively evaluated through field observations and empirical modeling \citep{barbour_2011,morris_2011,deleon_2013,johnston_2015,sandel_2015,usseglio_2017}. Determining the feasibility of such initiatives requires modeling the change in biomass in response to changes in fishing mortality (*i.e.* culling). A common way to model this is via length-structured population models, where fish lengths are converted to weight to calculate total biomass \citep{barbour_2011,cote_2014,andradibrown_2017}. The allometric length-weight relationship is thus an essential component of these models, but this relationship can vary across regions as a response to biotic and abiotic conditions \citep{johnson_2016}.

@@

@@

We obtained a total of 18 parameter pairs by combining length-weight parameters extracted from the literature and the additional pair calculated here (Fig \ref{fig:all_allo}). Recall that the objective of this study is not to describe variations between populations (which would require access to raw data), but rather to assess how *ex situ* parameter values influence the accuracy of weight estimates for lionfish, using the central Mexican Caribbean as a case study. Using each of the 18 parameter pairs, we estimated $TW$ from the $TL$ observations collected in the central Mexican Caribbean (n = 109, with $TL$ ranging from 34 mm to 310 mm) and divided predicted weights by known observed weights to obtain a simple measure of over- or underestimation. Difference in mean weight ratios were tested with an analysis of covariance (ANCOVA):

@@

Where $R_{ij}$ is the weight ratio for the $i$-th organism obtained with parameters from the $j$-th study, $\tilde{\mu}$ is a constant for all individuals, $a_j$ is the treatment effect (*i.e.* the difference induced by each study), $TL_{ij}$ is the covariate (*i.e.* Total Length) for the $i$-th subject in the $j$-th group with slope $\beta$, and $e_{ij}$ is the error term of the regression. Ratios were logit-transformed prior to analysis, and a *post-hoc* Tukey's test was used to identify groups where mean ratios did not differ. All analyses were performed in R version `r paste(version$major, version$minor, sep = ".")` \citep{rcore_2018}. Raw data and code used in this work are available on github at github.com/jcvdav/lionfish_biometry.

@@

ANCOVA results show that there were significant differences in our predicted weight ratios for the central Mexican Caribbean when using each of the different pairs of parameters ($F_{17, 1943} = 24.96; p < 0.001$; Fig \ref{fig:bio_ratio}). For example, the actual observed weights of the 109 lionfish from the Central Mexican Caribbean had a mean $\pm$ SD of `r filter(Predicted_biomass, Study_id == "Barbour") %$% mean(Weight) %>% formatC(digits = 2, format = "f")` $\pm$ `r filter(Predicted_biomass, Study_id == "Barbour") %$% sd(Weight) %>% formatC(digits = 2, format = "f")` g. However, if we used allometric parameter values from Banco Chinchorro in the Caribbean to predict weights from our observed length observations, we estimated a mean $\pm$ SD of `r filter(Predicted_biomass, Study_id == "Sabido2BC") %$% mean(pWeight) %>% formatC(digits = 2, format = "f")` $\pm$ `r filter(Predicted_biomass, Study_id == "Sabido2BC") %$% sd(pWeight) %>% formatC(digits = 2, format = "f")` g \citep{sabidoitz_2016}. If we similarly used parameter values from North Carolina in the Western Atlantic to estimate lionfish weights in the Central Mexican Caribbean, we found a mean $\pm$ SD `r filter(Predicted_biomass, Study_id == "Barbour") %$% mean(pWeight) %>% formatC(digits = 2, format = "f")` $\pm$ `r filter(Predicted_biomass, Study_id == "Barbour") %$% sd(pWeight) %>% formatC(digits = 2, format = "f")` g \citep{barbour_2011}. Weights predicted from these extreme parameters correspond to mean predicted-to-observed weight ratios of `r filter(Predicted_biomass, Study_id == "Sabido2BC") %$% mean(Biomass_ratio) %>% formatC(digits = 2, format = "f")` $\pm$ `r filter(Predicted_biomass, Study_id == "Sabido2BC") %$% sd(Biomass_ratio) %>% formatC(digits = 2, format = "f")` and `r filter(Predicted_biomass, Study_id == "Barbour") %$% mean(Biomass_ratio) %>% formatC(digits = 2, format = "f")` $\pm$ `r filter(Predicted_biomass, Study_id == "Barbour") %$% sd(Biomass_ratio) %>% formatC(digits = 2, format = "f")` (mean $\pm$ SD), respectively. Furthermore, for individual organisms, the largest differences between predicted and observed values resulted in ratios of `r min(Predicted_biomass$Biomass_ratio) %>% formatC(digits = 2, format = "f")` and `r max(Predicted_biomass$Biomass_ratio) %>% formatC(digits = 2, format = "f")` (*i.e.* the tails of each violin in Fig \ref{fig:bio_ratio}). If instead of looking at the mean ratios we looked at the total biomass (*i.e.* summing across organisms), these cases would lead to estimated total biomasses of 76.2% (4,363.53 g) and 140% (8,039.96 g) of true observed biomass (5,729.34 g). These were the two most extreme sets of parameters, but results consistently varie Use of *ex situ* parameters can result in erroneous weight and total biomass estimates for lionfish.

Tukey's *post-hoc* test suggested that weight ratios for the central Mexican Caribbean were not different from those obtained with parameters from Little Cayman, the Bahamas, and some sites in the Gulf of Mexico (Tukey's HSD $p > 0.05$). All weight estimates using parameters from the Gulf of Mexico and Western Atlantic were higher than observed values, and only parameters from the Caribbean produced weights smaller than observed (Fig \ref{fig:bio_ratio}). The regional average ($\pm$ SD) of predicted-to-observed weight ratios from these three regions were 1.24 $\pm$ 0.309, 1.41 $\pm$ 0.523, and 1.20 $\pm$ 0.423 for the Gulf of Mexico, Western Atlantic, and Caribbean, respectively. This suggests that the smallest errors are observed when using parameters from other locations in the Caribbean.

@@

```{r pred_obs, fig.height = 5.2, fig.width = 5.5, fig.cap = "\\label{fig:bio_ratio}Violin plot of predicted-to-observed weight ratios for 18 pairs of allometric parameters. Sex is indicated in parentheses. Blue circles indicate median values and like letters indicate values that do not differ significantly. For Sabido-Itza et al, 2016b, BC and PNAX make reference to Banco Chinchorro and Parque Nacional Arrecifes de Xcalak, two sites for which they report parameters."}
fig_bio_ratio
```

@@

Our results suggest that lionfish exhibit highly variable, spatially heterogeneous allometric relationships across their invaded range, and that this variation is relevant for managing invasions. Moreover, we show that the use of *ex situ* parameter values may lead to highly biased weight estimates. Our comparison of observed weights to those predicted with locally-informed parameters and *ex situ* parameters showed that weight can be overestimated by more than threefold, highlighting the need to use local information. Here we discuss the implications of our findings, possible shortcommings in our analyses, and highlight potential future research directions.

Differences in length-weight relationships have traditionally been highlighted as potential pitfalls to fishery management. For example, \citet{wilson_2012} show that small-scale variations in length-at-age and fishing mortality in other Scorpaeniformes translate to differential landings, effort, and catch per unit effort in the live fish fishery of California, and that these differences must be taken into account in management plans. The lionfish case poses the opposite scenario, where the manager desires to eradicate the species. To accurately gauge both the effectiveness of lionfish removal efforts and the resources needed to successfully manage an invasion, we must acknowledge and understand regional biological differences in important variables such as allometric growth parameters.

@@

We might be inclined to attribute all variation in the lionfish length-weight relationship to the spatial origin of these parameters. However, samples from the 12 studies included here were not only collected in different locations, but also at different points in time and across different depth and size ranges (See Supplementary Table 2 for an extended version of Table 1). While we could not evaluate how these factors influenced length-weight estimates from previous studies without raw data, we still show that a lack of locally-calculated parameters can induce significant bias when calculating weight from length observations. The magnitude of the bias discovered in this study and our lack of understanding the sources driving spatial variation for lionfish highlights the need to simultaneously collect length-weight information across the invaded range to test for spatially-induced patterns and to link these findings to previously suggested environmental and genetic structures. Such an endeavor would provide insight into lionfish biology and better inform management.

Applying parameter estimates to lengths outside the range of lengths originally used to estimate the parameters may also induce error. Our smallest observed organism was 34 mm in $TL$, and only two studies estimated parametrs with smaller organisms \citep{edwards_2014,sabidoitza_2016}. By contrast, our largest organism had a $TL$ of 310 mm, which is well within the range of all other studies (the next smallest maximum length was 325 mm; See Supplementary Table 2). Due to the power function describing the allometric relationship (*i.e.* Eq. 1), the error in weight estimates is larger when extrapolation is done for lengths that are larger than the maximum length used to estimate the parameters. Our estimates are therefore conservative because we only used parameter pairs from other studies to estimate weights for lionfish up to 310 mm in the Central Mexican Caribbean, well within the range of lengths for which other parameters were estimated.

>>>>

---
title: "Spatial variation in allometric growth of invasive lionfish has management implications"
author: "Juan Carlos Villaseñor-Derbez, Sean P. Fitzgerald"
institution: "Bren School of Environmental Sciences and Management, University of California Santa Barbara, Santa Barbara, California, U.S."
abstract: |
  Lionfish (\textit{Pterois volitans / miles}) are an invasive species in the Western Atlantic and the Caribbean. Improving management of invasive lionfish populations requires accurate total biomass estimates, which depend on accurate estimates of allometric growth, but sedentary species like lionfish often exhibit high levels of spatial variation in life history characteristics. We reviewed 17 published length-weight relationships for lionfish taken throughout their invasive range and found regional differences that led to significant misestimates when calculating weight from length observations. The spatial pattern we observed is consistent with findings from other studies focused on genetics or length-at-age. Here, the use of \textit{ex situ} parameter values resulted in total biomass estimates between 76.2% and 140% of true observed biomass, and up to a threefold under- or overestimation of total weight for an individual organism. These findings can have implications for management in terms of predicting effects on local ecosystems, evaluating the effectiveness of removal programs, or estimating biomass available for harvest.
keywords:
  - Lionfish
  - invasive species
  - length-weight
  - allometric growth
  - regional variations
bibliography: references.bib
output:
  pdf_document:
    keep_tex: yes
    number_sections: no
    template: peerj_template.tex
---

@@

The feasibility of establishing fisheries through lionfish removal programs has been extensively evaluated through field observations and empirical modeling \citep{barbour_2011,morris_2011,deleon_2013,johnston_2015,sandel_2015,usseglio_2017}. Determining the feasibility of such initiatives requires modeling the change in biomass in response to changes in fishing mortality (*i.e.,* culling). A common way to model this is via length-structured population models, where fish lengths are converted to weight to calculate total biomass \citep{barbour_2011,cote_2014,andradibrown_2017}. The allometric length-weight relationship is thus an essential component of these models, but this relationship can vary across regions as a response to biotic and abiotic conditions \citep{johnson_2016}.

@@

<!-- Is it possible to get rid of the double parentheses that appear in the .pdf? (e.g., line 75)  Or is this their format and they don't care and will fix it for us if/when it's published? -->

@@

We obtained a total of 18 parameter pairs by combining length-weight parameters extracted from the literature and the additional pair calculated here (Fig \ref{fig:all_allo}). Recall that the objective of this study is not to describe differences in the length-weight relationship between populations (which would require access to raw data), but rather to assess how *ex situ* parameter values influence the accuracy of weight estimates for lionfish, using the central Mexican Caribbean as a case study. Using each of the 18 parameter pairs, we estimated $TW$ from the $TL$ observations collected in the central Mexican Caribbean (n = 109, with $TL$ ranging from 34 mm to 310 mm) and divided predicted weights by known observed weights to obtain a simple measure of over- or underestimation. Difference in mean weight ratios were tested with an analysis of covariance (ANCOVA):

@@

Where $R_{ij}$ is the weight ratio for the $i$-th organism obtained with parameters from the $j$-th study, $\tilde{\mu}$ is a constant for all individuals, $a_j$ is the treatment effect (*i.e.,* the difference induced by each study), $TL_{ij}$ is the covariate (*i.e.,* Total Length) for the $i$-th subject in the $j$-th group with slope $\beta$, and $e_{ij}$ is the error term of the regression. Ratios were logit-transformed prior to analysis, and a *post-hoc* Tukey's test was used to identify groups where mean ratios did not differ. All analyses were performed in R version `r paste(version$major, version$minor, sep = ".")` \citep{rcore_2018}. Raw data and code used in this work are available on github at github.com/jcvdav/lionfish_biometry.

@@

ANCOVA results revealed significant differences in our predicted weight ratios for the central Mexican Caribbean when using each of the different pairs of parameters ($F_{17, 1943} = 24.96; p < 0.001$; Fig \ref{fig:bio_ratio}). For example, the actual observed weights of the 109 lionfish from the Central Mexican Caribbean had a mean $\pm$ SD of `r filter(Predicted_biomass, Study_id == "Barbour") %$% mean(Weight) %>% formatC(digits = 2, format = "f")` $\pm$ `r filter(Predicted_biomass, Study_id == "Barbour") %$% sd(Weight) %>% formatC(digits = 2, format = "f")` g. However, if we used allometric parameter values from Banco Chinchorro in the Caribbean to predict weights from our observed length observations, we estimated a mean $\pm$ SD of `r filter(Predicted_biomass, Study_id == "Sabido2BC") %$% mean(pWeight) %>% formatC(digits = 2, format = "f")` $\pm$ `r filter(Predicted_biomass, Study_id == "Sabido2BC") %$% sd(pWeight) %>% formatC(digits = 2, format = "f")` g \citep{sabidoitz_2016}. If we similarly used parameter values from North Carolina in the Western Atlantic to estimate lionfish weights in the Central Mexican Caribbean, we found a mean $\pm$ SD `r filter(Predicted_biomass, Study_id == "Barbour") %$% mean(pWeight) %>% formatC(digits = 2, format = "f")` $\pm$ `r filter(Predicted_biomass, Study_id == "Barbour") %$% sd(pWeight) %>% formatC(digits = 2, format = "f")` g \citep{barbour_2011}. Weights predicted from these extreme parameters correspond to mean predicted-to-observed weight ratios of `r filter(Predicted_biomass, Study_id == "Sabido2BC") %$% mean(Biomass_ratio) %>% formatC(digits = 2, format = "f")` $\pm$ `r filter(Predicted_biomass, Study_id == "Sabido2BC") %$% sd(Biomass_ratio) %>% formatC(digits = 2, format = "f")` and `r filter(Predicted_biomass, Study_id == "Barbour") %$% mean(Biomass_ratio) %>% formatC(digits = 2, format = "f")` $\pm$ `r filter(Predicted_biomass, Study_id == "Barbour") %$% sd(Biomass_ratio) %>% formatC(digits = 2, format = "f")` (mean $\pm$ SD), respectively. Furthermore, for the smallest and largest individual organisms collected in the central Mexican Caribbean (TL = 34 and 310 mm, respectively), the differences between predicted and observed values resulted in ratios of `r min(Predicted_biomass$Biomass_ratio) %>% formatC(digits = 2, format = "f")` and `r max(Predicted_biomass$Biomass_ratio) %>% formatC(digits = 2, format = "f")` (*i.e.,* the tails of each violin in Fig \ref{fig:bio_ratio}). If we examined biomass (*i.e.,* summing across all 109 organisms) instead of mean ratios, total biomass estimates were 76.2% (4,363.53 g) and 140% (8,039.96 g) of true observed biomass (5,729.34 g). We produced these misestimates using the two most extreme sets of parameters, but results varied consistently across locations. Overall, the use of *ex situ* parameters led to significantly erroneous estimates of individual weight and total biomass for lionfish.

Tukey's *post-hoc* test showed that weight ratios for the central Mexican Caribbean were not different from those obtained with parameters from Little Cayman, the Bahamas, and some sites in the Gulf of Mexico (Tukey's HSD $p > 0.05$). All weight estimates using parameters from the Gulf of Mexico and Western Atlantic were higher than observed values, and only parameters from the Caribbean produced weights smaller than observed (Fig \ref{fig:bio_ratio}). The regional average ($\pm$ SD) of predicted-to-observed weight ratios from these three regions were 1.24 $\pm$ 0.309, 1.41 $\pm$ 0.523, and 1.20 $\pm$ 0.423 for the Gulf of Mexico, Western Atlantic, and Caribbean, respectively. This suggests that the smallest errors are observed when using parameters from other locations in the Caribbean.

<!-- One minor suggestion - instead of leading with where the differences AREN'T, start with where they ARE - e.g., "Tukey's *post-hoc* test showed that weight ratios for the central Mexican Caribbean differeed from those obtained with parameters from the Western Atlantic, and most sites in the Caribbean and the Gulf of Mexico (Tukey's HSD $p > 0.05$). The only sites where weight ratios did not differ from the central Mexican Caribbean were Little Cayman, the Bahamas, and some sites (WHICH ONES?) in the Gulf of Mexico (Tukey's HSD $p > 0.05$)." -->

@@

```{r pred_obs, fig.height = 5.2, fig.width = 5.5, fig.cap = "\\label{fig:bio_ratio}Violin plot of predicted-to-observed weight ratios when applying each of 18 different pairs of allometric parameters to the 109 lionfish collected in the central Mexican Caribbean. Sex is indicated in parentheses. Blue circles indicate median values and like letters indicate values that do not differ significantly. For Sabido-Itza et al, 2016b, BC and PNAX make reference to Banco Chinchorro and Parque Nacional Arrecifes de Xcalak, two sites for which they report parameters."}
fig_bio_ratio
```

@@

Our results suggest that lionfish exhibit highly variable, spatially heterogeneous allometric relationships across their invaded range, and that this variation is relevant for managing invasions. Moreover, we show that the use of *ex situ* parameter values may lead to highly biased weight estimates. Our comparison of observed weights to those predicted with locally-informed parameters and *ex situ* parameters showed that weight of an individual lionfish can be overestimated by more than threefold, highlighting the need to use local information. Here we discuss the implications of our findings, possible shortcommings in our analyses, and highlight potential future research directions.

Differences in length-weight relationships have traditionally been highlighted as potential pitfalls to fishery management. For example, \citet{wilson_2012} showed that small-scale variations in length-at-age and fishing mortality in other Scorpaeniformes translate to differential landings, effort, and catch per unit effort in the live fish fishery of California, and that these differences must be taken into account in management plans. The lionfish case poses the opposite scenario, where the manager desires to eradicate the species. To accurately gauge both the effectiveness of lionfish removal efforts and the resources needed to successfully manage an invasion, we must acknowledge and understand regional biological differences in important variables such as allometric growth parameters.

@@

We might be inclined to attribute all variation in the lionfish length-weight relationship to the spatial origin of these parameters. However, samples from the 12 studies included here were not only collected in different locations, but also at different points in time and across different depth and size ranges (See Supplementary Table 2 for an extended version of Table 1). The magnitude of the bias discovered in this study and our lack of understanding the sources driving spatial variation for lionfish highlights the need to simultaneously collect length-weight information across the invaded range to test for spatially-induced patterns and to link these findings to previously suggested environmental and genetic structures. Such an endeavor would provide insight into lionfish biology and better inform management. However, while we could not evaluate how these factors influenced length-weight estimates from previous studies without raw data, we still show that a lack of locally-calculated parameters can induce significant bias when calculating weight from length observations. We demonstrate the importance of using *in situ* parameters to obtain accurate weight estimates regardless of the underlying mechanisms driving variation between populations.

Applying parameter estimates to lengths outside the range of lengths originally used to estimate the parameters may also induce error. Our smallest observed organism was 34 mm in $TL$, and only two studies estimated parametrs with smaller organisms \citep{edwards_2014,sabidoitza_2016}. By contrast, our largest organism had a $TL$ of 310 mm, which is well within the range of all other studies (the next smallest maximum length was 325 mm; See Supplementary Table 2). Due to the power function describing the allometric relationship (*i.e.,* Eq. 1), the error in weight estimates is larger when extrapolation is done for lengths that are larger than the maximum length used to estimate the parameters. Our estimates are therefore conservative because we only used parameter pairs from other studies to estimate weights for lionfish up to 310 mm in the Central Mexican Caribbean, well within the range of lengths for which other parameters were estimated.
